---
title: "R Basics for Psychological Research"
subtitle: "Descriptives, Correlation, and Multiple Regression"
author: "Dr. Anisha Singh"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    embed-resources: true
---

## Overview

In this session we will work through a complete data analysis workflow using a simulated dataset from psychological research. The dataset contains the following variables for **150 students**:

| Variable | Description | Scale |
|---|---|---|
| `student_id` | Participant identifier | ID only |
| `study_hours` | Average daily study hours | 0-12 hours |
| `sleep_hours` | Average nightly sleep | 3-11 hours |
| `anxiety_score` | Trait anxiety (adapted from STAI) | 20-80 (higher = more anxiety) |
| `exam_score` | Final exam performance | 0-100 |

**Research question:** Do study hours, sleep hours, and anxiety predict exam performance?

---

## Step 1: Load Libraries

Use `install.packages("psych")` once if you have not installed it yet. Then load it each session:

```{r}
#| label: load-libraries
#| message: false
#| warning: false

library(psych)   # for the describe() function
```

A library (package) is a collection of pre-written functions that extend base R. Load with `library()` at the top of every script.

---

## Step 2: Import the Dataset

```{r}
#| label: import-data

df <- read.csv("psych_study_data.csv")
head(df)     # first 6 rows
dim(df)      # rows x columns
names(df)    # variable names
```

Always inspect your data after importing to confirm row counts and variable names look correct.

---

## Step 3: Descriptive Statistics

### 3a. Mean and Standard Deviation (base R)

```{r}
#| label: means

mean(df$exam_score)
mean(df$study_hours)
mean(df$sleep_hours)
mean(df$anxiety_score)
```

```{r}
#| label: sds

sd(df$exam_score)
sd(df$study_hours)
sd(df$sleep_hours)
sd(df$anxiety_score)
```

A larger SD indicates more variability around the mean; a smaller SD means scores cluster tightly.

### 3b. Rich Descriptives with psych::describe()

```{r}
#| label: describe

# Drop column 1 (student_id) before describing
describe(df[, -1])
```

Key columns: n (sample size), mean, sd, median, min, max, skew (0 = symmetric), kurtosis (0 = normal), se (standard error of the mean).

---

## Step 4: Correlation with cor.test()

`cor.test()` returns Pearson r and tests whether it differs significantly from zero.

### Study Hours and Exam Score

```{r}
#| label: cor-study

cor.test(df$study_hours, df$exam_score)
```

### Anxiety Score and Exam Score

```{r}
#| label: cor-anxiety

cor.test(df$anxiety_score, df$exam_score)
```

### Sleep Hours and Exam Score

```{r}
#| label: cor-sleep

cor.test(df$sleep_hours, df$exam_score)
```

Reading the output:
- cor: Pearson r (range -1 to +1)
- t: test statistic
- df: degrees of freedom (n - 2)
- p-value: probability of this result if the true r = 0
- 95 percent confidence interval: plausible range for the true r

Effect size convention (Cohen, 1988): |r| < .10 = negligible; .10-.29 = small; .30-.49 = moderate; >= .50 = large.

---

## Step 5: Multiple Regression

Multiple regression estimates how several predictors jointly explain an outcome and the unique contribution of each predictor controlling for the others.

```{r}
#| label: regression

model <- lm(exam_score ~ study_hours + sleep_hours + anxiety_score, data = df)
summary(model)
```

Reading the output:

Coefficients table:
- Estimate: regression slope (B)
- Std. Error: precision of the estimate
- t value: test statistic (H0: B = 0)
- Pr(>|t|): p-value; *** <= .001, ** <= .01, * <= .05

Model fit:
- Multiple R-squared: proportion of variance in exam score explained by all predictors combined
- Adjusted R-squared: R-squared penalized for number of predictors (preferred for comparisons)
- F-statistic: tests whether the overall model is significant

Interpreting a slope: for every one-unit increase in a predictor, exam score changes by B points holding all other predictors constant.

---

## Step 6: Scatterplot (Bonus)

```{r}
#| label: plot
#| fig-width: 6
#| fig-height: 4

plot(df$study_hours, df$exam_score,
     xlab = "Study Hours (per day)",
     ylab = "Exam Score",
     main = "Study Hours and Exam Performance",
     pch = 19, col = "steelblue", cex = 0.7)
abline(lm(exam_score ~ study_hours, data = df),
       col = "firebrick", lwd = 2)
```

---

## Function Quick Reference

```{r}
#| eval: false

library(psych)
df <- read.csv("psych_study_data.csv")
head(df)
dim(df)
names(df)
mean(df$variable)
sd(df$variable)
describe(df[, -1])
cor.test(df$x, df$y)
model <- lm(y ~ x1 + x2 + x3, data = df)
summary(model)
```

---

## References

Cohen, J. (1988). Statistical power analysis for the behavioral sciences (2nd ed.). Lawrence Erlbaum Associates.

Revelle, W. (2023). psych: Procedures for psychological, psychometric, and personality research. R package version 2.3.6.
